---
title: "Cyclistic - Case Study 1"
author: "Saturnino"
date: "2022-12-01"
output: html_document
---

This is an analysis on data from 2021-11 until 2022-10 on bike usage by Cyclistic users.

## Processing data

The datasets have been inspected and modified in excel prior to being loaded in R.

### Documentation of cleaning and manipulation before loading data in R

##### Added a “ride_length”, “day_of_week”, and “month” column to all .csv files.
* “ride_length” is equal to “ended_at” minus “started_at” column. (=”ended_at”-“started_at”)
* “day_of_week” created using the WEEKDAY function on column “started_at”. (=WEEKDAY(“started_at”,2)
* “month” created using the MONTH function on column “started_at”. (=MONTH(“started_at”)
* “ride_length” formatted as time, HH:MM:SS format.
* “day_of_week” and “month” formatted as number with no decimals.

##### Removing duplicates using the “Remove Duplicates” function in column “ride_id”.

##### Checking for errors and blanks using filters for all columns.

##### Not removing blanks or adding data to cells in columns “start_station_name”, “start_station_id”, “end_station_name”, “end_station_id”, “start_lat”, “start_lng”, “end_lat”, “end_lng”.

##### Data for 2021-11
* Removed 12 duplicates.
*	Removed 53 entries where “ended_at” was before “started_at”, resulting in an error for “ride_length”. All errors are from the date 2021-11-07.
*	Removed blank rows.

##### Data for 2021-12
*	Clean, no duplicates, blank rows, or errors.

##### Data for 2022-01
*	Clean, no duplicates, blank rows, or errors.

##### Data for 2022-02
*	Clean, no duplicates, blank rows, or errors.

##### Data for 2022-03
*	Removed 2 entries where “ended_at” was before “started_at”, resulting in an error for “ride_length”. Both errors are from the date 2022-03-05.
*	No duplicates or blank rows.

##### Data for 2022-04
*	Clean, no duplicates, blank rows, or errors.

##### Data for 2022-05
*	Clean, no duplicates, blank rows, or errors.

##### Data for 2022-06
*	Removed 1 duplicate.
*	No blank rows or errors.

##### Data for 2022-07
*	Removed 1 duplicate.
*	Removed blank rows.
*	No errors.

##### Data for 2022-08
*	Clean, no duplicates, blank rows, or errors.

##### Data for 2022-09
*	Clean, no duplicates, blank rows, or errors.

##### Data for 2022-10
*	Removed 4 entries where “ended_at” was before “started_at”, resulting in an error for “ride_length”. Errors dated: 2022-10-03, 2022-10-13, 2022-10-21, 2022-10-24.
*	No duplicates or blank rows.

### Loadint data into R

```{r Loading data}
#Loading libraries
library(tidyverse)
library(ggplot2)
library(lubridate)

#Setting current folder to dataset location
setwd("F:/case_study_1_cyclistic")

#Loading .csv files
data_21_11 <- read.csv("202111-divvy-tripdata.csv")
data_21_12 <- read.csv("202112-divvy-tripdata.csv")
data_22_01 <- read.csv("202201-divvy-tripdata.csv")
data_22_02 <- read.csv("202202-divvy-tripdata.csv")
data_22_03 <- read.csv("202203-divvy-tripdata.csv")
data_22_04 <- read.csv("202204-divvy-tripdata.csv")
data_22_05 <- read.csv("202205-divvy-tripdata.csv")
data_22_06 <- read.csv("202206-divvy-tripdata.csv")
data_22_07 <- read.csv("202207-divvy-tripdata.csv")
data_22_08 <- read.csv("202208-divvy-tripdata.csv")
data_22_09 <- read.csv("202209-divvy-publictripdata.csv")
data_22_10 <- read.csv("202210-divvy-tripdata.csv")
```

### Data manipulation in R

##### After loading the data we combine it into one dataset "data_year".

```{r Creating dataset}
#Combining all months into one dataset
data_year <- bind_rows(data_21_11, data_21_12, data_22_01, data_22_02, data_22_03, data_22_04, data_22_05, data_22_06, data_22_07, data_22_08, data_22_09, data_22_10)
```

##### Calculating ride length into seconds for easier calculation later, adding results into a new column "ride_length_2".

```{r Ride length calculation}
#Calculating ride length as seconds
data_year$ride_length_2 <- difftime(as_datetime(data_year$ended_at,format=("%d-%m-%Y %H:%M")), as_datetime(data_year$started_at,format=("%d-%m-%Y %H:%M")))

```

##### Changing the data type of "ride_length_2".

```{r Data type change}
#Converting ride_length_2 as numeric value
data_year$ride_length_2<- as.numeric(as.character(data_year$ride_length_2))
```

##### Removing negatives in "ride_length_2" and creating a new version of the dataset "data_year_v2".

```{r Removing negatives - new dataset}
#Removing negatives missed in "ride_length_2" and creating a new version of the dataset
data_year_v2<-data_year[!(data_year$ride_length_2<0),]
```

##### Inspection of data before analysis.

```{r Inspecting data}
#Inspecting data before analysis
colnames(data_year_v2)
nrow(data_year_v2)
dim(data_year_v2)
head(data_year_v2)
str(data_year_v2)
summary(data_year_v2)
```

## Data analysis

##### Descriptive analysis on ride length.

```{r Ride length analysis}
#Descriptive analysis on ride length
seconds_to_period(mean(data_year_v2$ride_length_2))
seconds_to_period(median(data_year_v2$ride_length_2))
seconds_to_period(max(data_year_v2$ride_length_2))
seconds_to_period(min(data_year_v2$ride_length_2))
```

##### Comparing member and casual ride lengths.

```{r Member vs casual}
#Comparing member and casual
aggregate(data_year_v2$ride_length_2 ~ data_year_v2$member_casual, FUN = mean)
aggregate(data_year_v2$ride_length_2 ~ data_year_v2$member_casual, FUN = median)
aggregate(data_year_v2$ride_length_2 ~ data_year_v2$member_casual, FUN = max)
aggregate(data_year_v2$ride_length_2 ~ data_year_v2$member_casual, FUN = min)
```

##### Comparing member and casual ride lengths by weekday.

```{r Weekday comparison}
#Comparing by day
aggregate(data_year_v2$ride_length_2 ~ data_year_v2$member_casual + data_year_v2$day_of_week, FUN = mean)
aggregate(data_year_v2$ride_length_2 ~ data_year_v2$member_casual + data_year_v2$day_of_week, FUN = median)
```

##### Comparing member and casual ride lengths by month.

```{r Month comparison}
#Comparing by month
aggregate(data_year_v2$ride_length_2 ~ data_year_v2$member_casual + data_year_v2$month, FUN = mean)
aggregate(data_year_v2$ride_length_2 ~ data_year_v2$member_casual + data_year_v2$month, FUN = median)
```

##### Member and casual use of rideable types.

```{r Comparing rideables}
#Rideable types compared
data_year_v2 %>% count(rideable_type, member_casual)
```

##### Comparing member and casual ride counts and ride lengths by day.

```{r Ride count by day}
#Ridership by day and membership
data_year_v2 %>%
  mutate(weekday=wday(started_at,label=TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(num_of_rides=n(), average_duration=seconds_to_period(mean(ride_length_2)), median_ride_length=seconds_to_period(median(ride_length_2))) %>%
  arrange(member_casual, weekday)
```

##### Comparing member and casual ride counts and ride lengths by month.

```{r Ride count by month}
#Ridership by month and membership
data_year_v2 %>%
  group_by(member_casual,month) %>% 
  summarise(num_of_rides=n(), average_duration=seconds_to_period(mean(ride_length_2)), median_ride_length=seconds_to_period(median(ride_length_2))) %>%
  arrange(member_casual, month)
```

##### Visual comparing member and casual ride counts by day.

```{r Visual by day}
#Visual for ridership by day
data_year_v2 %>%
  mutate(weekday=wday(started_at,label=TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(num_of_rides=n()) %>%
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = num_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

##### Visual comparing member and casual ride counts by month.

```{r Visual by month}
#visual for ridership by month
data_year_v2 %>%
  group_by(member_casual,month) %>% 
  summarise(num_of_rides=n()) %>%
  arrange(member_casual, month) %>% 
  ggplot(aes(x = month, y = num_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

##### Visual comparing member and casual daily ride length averages.

```{r Visual daily ride length}
#Visual for ridership day averages
data_year_v2 %>%
  mutate(weekday=wday(started_at,label=TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(num_of_rides=n(), average_duration=mean(ride_length_2)) %>%
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

##### Visual comparing member and casual monthly ride length averages.

```{r Visual monthly ride length}
#Visual for ridership month averages
data_year_v2 %>%
  group_by(member_casual,month) %>% 
  summarise(num_of_rides=n(), average_duration=mean(ride_length_2)) %>%
  arrange(member_casual, month) %>% 
  ggplot(aes(x = month, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

##### Visual comparing member and casual daily ride medians.

```{r Visual daily median}
#Visual for ridership day medians
data_year_v2 %>%
  mutate(weekday=wday(started_at,label=TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(num_of_rides=n(), median_ride_length=median(ride_length_2)) %>%
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = median_ride_length, fill = member_casual)) +
  geom_col(position = "dodge")
```

##### Visual comparing member and casual monthly ride medians.

```{r Visual monthly medians}
#Visual for ridership month medians
data_year_v2 %>%
  group_by(member_casual,month) %>% 
  summarise(num_of_rides=n(), median_ride_length=median(ride_length_2)) %>%
  arrange(member_casual, month) %>% 
  ggplot(aes(x = month, y = median_ride_length, fill = member_casual)) +
  geom_col(position = "dodge")
```

## Exporting data

##### Exporting data for further analysis and visualization.

```{r Export}
#.csv data for export
daily_avg_med<-data_year_v2 %>%
  mutate(weekday=wday(started_at,label=TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(num_of_rides=n(), average_duration=seconds_to_period(mean(ride_length_2)), median_ride_length=seconds_to_period(median(ride_length_2))) %>%
  arrange(member_casual, weekday)
write.csv(daily_avg_med, file="daily_avg_med.csv")

monthly_avg_med<-data_year_v2 %>%
  group_by(member_casual,month) %>% 
  summarise(num_of_rides=n(), average_duration=seconds_to_period(mean(ride_length_2)), median_ride_length=seconds_to_period(median(ride_length_2))) %>%
  arrange(member_casual, month)
write.csv(monthly_avg_med, file="monthly_avg_med.csv")

rideable_comp<-data_year_v2 %>% count(rideable_type, member_casual)
write.csv(rideable_comp, file="rideable_comp.csv")
```